Code Club Project: Assessing whether intra and inter-genomic variation hinder utility of ESVs and ASVs

### 2021-01-12 submission/manuscript.Rmd
Format inline code

```{r, echo=FALSE}
library(tidyverse)
library(knitr)
library(here)
```

@@

16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s. The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences. ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold. However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical. Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU. An example of both is seen in the comparison of Staphylococcus aureus (NCTC 8325) and S. epidermidis (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs. Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU. The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.

```{r, echo=FALSE}
metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character()))

@@

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g. `r single_copy$species`) and `r max_copy$median` (e.g. `r max_copy$species`) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**). On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**). For example, `r single_copy$species` generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a E. coli genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` E. coli genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

@@

To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds. I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species. By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC). Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU. First, I identified the thresholds where the sensitivity and specificity were most similar to each other. For this criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and XX (V4V5). Second, I identified the threshold that resulted in the point on the ROC curve that was closest to perfect classification. For this criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and XX (V4V5). Surprisingly, these analyses revealed that thresholds near 3% balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.

>>>>

```{r, echo=FALSE}
library(tidyverse)
library(knitr)
library(here)

opts_chunk$set(echo = FALSE)

inline_hook <- function(x){

	if(is.numeric(x)){
		if(abs(x - round(x) < .Machine$double.eps)){
			# treat as integer
			formatted <- format(x, digits=0, big.mark = ",")

		} else {
			# treat as floating point number
			formatted <- format(x, digits=2, nsmall=2, big.mark = ",")
		}
	} else {
		formatted <-x
	}

	paste0("**", formatted, "**")

}

knit_hooks$set(inline=inline_hook)

format_pct <- function(x, my_digits=1){

	format(x, digits=my_digits, nsmall=my_digits)

}
```

@@

16S rRNA gene sequencing is a very powerful technique for describing and comparing microbial communities. Efforts to link 16S rRNA gene sequences to taxonomic levels based on distance thresholds go back to at least the 1990s. The distance-based thresholds that were developed and are now widely used (3%) were based on DNA-DNA hybridization approaches that are not as precise as genome sequencing. Instead, genome sequencing technologies have suggested that the widely used 3% distance threshold to operationally define bacterial taxa is too coarse. As an alternative to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way to adopt the thresholds suggested by genome sequencing to microbial community analysis using 16S rRNA gene sequences. ASVs are a unit of microbial community inference that do not cluster sequences based on a distance-based threshold. However, most bacterial genomes have more than 1 copy of the rrn operon and those copies are not identical. Therefore, using too fine a threshold to identify OTUs creates the risk of splitting a single genome into multiple bins and using too broad of a threshold to define OTUs creates the risk of lumping together bacterial species into the same OTU. An example of both is seen in the comparison of *Staphylococcus aureus* (NCTC 8325) and *S. epidermidis* (ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10 copies of the 16S rRNA gene each have a different sequence and so if OTUs are defined based on ASVs, each genome would be split into 5 OTUs. Conversely, if the copies were clustered using a 3% distance threshold all 10 copies would cluster into the same OTU. The goal of this study was to quantify the risk of splitting a single genome into multiple bins and the risk of lumping together different bacterial species into the same bin.

```{r}
metadata <- read_tsv(here("data/references/genome_id_taxonomy.tsv"),
										 col_types = cols(.default = col_character()))

@@

To investigate the variation in the number of copies of the 16S rRNA gene per genome as well as the intragenomic variation among copies of the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn copy number database (rrnDB; CITATION). Among the `r n_distinct(metadata$species)` species represented in the rrnDB there were `r nrow(metadata)` genomes. The median number of rrn operson per species ranged between `r single_copy$median` (e.g. *`r single_copy$species`*) and `r max_copy$median` (*`r max_copy$species`*) copies of the rrn operon. As the number of copies of the operon in a genome increased, the number of variants of the 16S rRNA gene in each genome also increased (**FIGURE**). On average, there were `r rate_v19` variants per copy of the full length 16S rRNA gene and an average of `r rate_v4`, `r rate_v34`, and `r rate_v45` variants when considering the V4, V3-V4, and V4-V5 regions of the gene, respectively. Although a species tended to have a consistent number of 16S rRNA gene copies per genome, the number of total variants increased with the number of genomes that were sampled (**FIGURE**). For example, *`r single_copy$species`* generally only had `r single_copy$median` copy of the gene per genome, but across the `r single_copy_N` genomes that have been sequenced there were `r single_nasvs` versions of the gene. Similarly, a *E. coli* genome typically had `r ecoli_copy$median` copies of the 16S rRNA gene with between `r ecoli_copy$min` and `r ecoli_copy$max` distinct full length sequences per genome. Across the `r ecoli_copy$N` *E. coli* genomes that have been sequenced, there were `r ecoli_nasvs` different variants of the gene. These observations highlight the risk of selecting a threshold for defining units of inference that is too narrow because it is possible to split a single genome into multiple units.

@@

```{r}
sensitivity_specificity <- read_tsv(here("data/processed/rrnDB.roc.tsv"),
													 col_types = cols(.default = col_double(),
													 								 region = col_character()
													 								 )
													 ) %>%
	mutate(sensitivity = true_pos / (true_pos + false_neg),
				 specificity = true_neg / (true_neg + false_pos)) %>%
	group_by(region, threshold) %>%
	summarize(sensitivity = median(sensitivity),
						specificity = median(specificity),
						.groups="drop") %>%
	select(region, threshold, sensitivity, specificity)

balance <- sensitivity_specificity %>%
	mutate(diff=abs(sensitivity - specificity)) %>%
	group_by(region) %>%
	summarize(min_diff = min(diff),
						threshold = threshold[which.min(diff)],
						sensitivity = sensitivity[which.min(diff)],
						specificity = specificity[which.min(diff)],
						.groups="drop"
						)

balance_v19 <- balance %>% filter(region == "v19") %>% pull(threshold) * 100
balance_v4 <- balance %>% filter(region == "v4") %>% pull(threshold) * 100
balance_v34 <- balance %>% filter(region == "v34") %>% pull(threshold) * 100
balance_v45 <- balance %>% filter(region == "v45") %>% pull(threshold) * 100

distance <- sensitivity_specificity %>%
	mutate(distance = sqrt((specificity - 1)^2 + (sensitivity - 1)^2)) %>%
	group_by(region) %>%
	summarize(min_distance = min(distance),
						threshold = threshold[which.min(distance)],
						sensitivity = sensitivity[which.min(distance)],
						specificity = specificity[which.min(distance)],
						.groups="drop"
						)

distance_v19 <- distance %>% filter(region == "v19") %>% pull(threshold) * 100
distance_v4 <- distance %>% filter(region == "v4") %>% pull(threshold) * 100
distance_v34 <- distance %>% filter(region == "v34") %>% pull(threshold) * 100
distance_v45 <- distance %>% filter(region == "v45") %>% pull(threshold) * 100

```

To provide a more nuanced approach to selecting a threshold, it would be useful to to quantify the sensitivity and specificity of characterizing bacterial species using OTUs defined at different thresholds. I created confusion matrices for multiple regions of the 16S rRNA gene: true positives were those cases where two ASVs were joined in the same OTU and the same species; true negatives were those cases where two ASVs from different OTUs came from different species; false positives were those ASVs that joined the same OTU, but were from different species; and false negatives were those ASVs that joined different OTUs, but were from the same species. By calculating the sensitivity and specificity for each threshold and each region of the 16S rRNA gene, I was able to constuct a receiver operator characteristic curve (ROC). Because the ROC curve represents a range of possible thresholds and sensitivities and specificities, I used two metrics to select the best threshold for defining an OTU. First, I identified the thresholds where the sensitivity and specificity were most similar to each other. For this criterion, the best distance thresholds were `r format_pct(balance_v19)`% (V1-V9), `r format_pct(balance_v4)`% (V4), `r format_pct(balance_v34)`% (V3-V4), and `r format_pct(balance_v45)`% (V4-V5). Second, I identified the distance threshold that resulted in the point on the ROC curve that was closest to perfect classification. For this criterion, the best distance thresholds were `r format_pct(distance_v19)`% (V1-V9), `r format_pct(distance_v4)`% (V4), `r format_pct(distance_v34)`% (V3-V4), and `r format_pct(distance_v45)`% (V4-V5). Surprisingly, these analyses revealed that thresholds near 3% distance balance the risks of splitting genomes into separate OTUs and lumping species into the same OTU.

### 2021-01-12 submission/manuscript.tex
Format inline code

16S rRNA gene sequencing is a very powerful technique for describing and
comparing microbial communities. Efforts to link 16S rRNA gene sequences
to taxonomic levels based on distance thresholds go back to at least the
1990s. The distance-based thresholds that were developed and are now
widely used (3\%) were based on DNA-DNA hybridization approaches that
are not as precise as genome sequencing. Instead, genome sequencing
technologies have suggested that the widely used 3\% distance threshold
to operationally define bacterial taxa is too coarse. As an alternative
to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way
to adopt the thresholds suggested by genome sequencing to microbial
community analysis using 16S rRNA gene sequences. ASVs are a unit of
microbial community inference that do not cluster sequences based on a
distance-based threshold. However, most bacterial genomes have more than
1 copy of the rrn operon and those copies are not identical. Therefore,
using too fine a threshold to identify OTUs creates the risk of
splitting a single genome into multiple bins and using too broad of a
threshold to define OTUs creates the risk of lumping together bacterial
species into the same OTU. An example of both is seen in the comparison
of Staphylococcus aureus (NCTC 8325) and S. epidermidis (ATCC 12228)
where each genome has 5 copies of the 16S rRNA gene. The 10 copies of
the 16S rRNA gene each have a different sequence and so if OTUs are
defined based on ASVs, each genome would be split into 5 OTUs.
Conversely, if the copies were clustered using a 3\% distance threshold
all 10 copies would cluster into the same OTU. The goal of this study
was to quantify the risk of splitting a single genome into multiple bins
and the risk of lumping together different bacterial species into the
same bin.

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB; CITATION). Among the 4774 species
represented in the rrnDB there were 15614 genomes. The median number of
rrn operson per species ranged between 1 (e.g.~Mycobacterium
tuberculosis) and 19 (e.g.~Metabacillus litoralis) copies of the rrn
operon. As the number of copies of the operon in a genome increased, the
number of variants of the 16S rRNA gene in each genome also increased
(\textbf{FIGURE}). On average, there were 0.5970303 variants per copy of
the full length 16S rRNA gene and an average of 0.2551013, 0.3265188,
and 0.2744228 variants when considering the V4, V3-V4, and V4-V5 regions
of the gene, respectively. Although a species tended to have a
consistent number of 16S rRNA gene copies per genome, the number of
total variants increased with the number of genomes that were sampled
(\textbf{FIGURE}). For example, Mycobacterium tuberculosis generally
only had 1 copy of the gene per genome, but across the 180 genomes that
have been sequenced there were 11 versions of the gene. Similarly, a E.
coli genome typically had 7 copies of the 16S rRNA gene with between 6
and 10 distinct full length sequences per genome. Across the 958 E. coli
genomes that have been sequenced, there were 1013 different variants of
the gene. These observations highlight the risk of selecting a threshold
for defining units of inference that is too narrow because it is
possible to split a single genome into multiple units.

@@

To provide a more nuanced approach to selecting a threshold, it would be
useful to to quantify the sensitivity and specificity of characterizing
bacterial species using OTUs defined at different thresholds. I created
confusion matrices for multiple regions of the 16S rRNA gene: true
positives were those cases where two ASVs were joined in the same OTU
and the same species; true negatives were those cases where two ASVs
from different OTUs came from different species; false positives were
those ASVs that joined the same OTU, but were from different species;
and false negatives were those ASVs that joined different OTUs, but were
from the same species. By calculating the sensitivity and specificity
for each threshold and each region of the 16S rRNA gene, I was able to
constuct a receiver operator characteristic curve (ROC). Because the ROC
curve represents a range of possible thresholds and sensitivities and
specificities, I used two metrics to select the best threshold for
defining an OTU. First, I identified the thresholds where the
sensitivity and specificity were most similar to each other. For this
criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
XX (V4V5). Second, I identified the threshold that resulted in the point
on the ROC curve that was closest to perfect classification. For this
criterion, the best thresholds were XX (V1-V9), XX (V4), XX (V3V4), and
XX (V4V5). Surprisingly, these analyses revealed that thresholds near
3\% balance the risks of splitting genomes into separate OTUs and
lumping species into the same OTU.

>>>>

16S rRNA gene sequencing is a very powerful technique for describing and
comparing microbial communities. Efforts to link 16S rRNA gene sequences
to taxonomic levels based on distance thresholds go back to at least the
1990s. The distance-based thresholds that were developed and are now
widely used (3\%) were based on DNA-DNA hybridization approaches that
are not as precise as genome sequencing. Instead, genome sequencing
technologies have suggested that the widely used 3\% distance threshold
to operationally define bacterial taxa is too coarse. As an alternative
to OTUs, amplicon sequencing variants (ASVs) have been proposed as a way
to adopt the thresholds suggested by genome sequencing to microbial
community analysis using 16S rRNA gene sequences. ASVs are a unit of
microbial community inference that do not cluster sequences based on a
distance-based threshold. However, most bacterial genomes have more than
1 copy of the rrn operon and those copies are not identical. Therefore,
using too fine a threshold to identify OTUs creates the risk of
splitting a single genome into multiple bins and using too broad of a
threshold to define OTUs creates the risk of lumping together bacterial
species into the same OTU. An example of both is seen in the comparison
of \emph{Staphylococcus aureus} (NCTC 8325) and \emph{S. epidermidis}
(ATCC 12228) where each genome has 5 copies of the 16S rRNA gene. The 10
copies of the 16S rRNA gene each have a different sequence and so if
OTUs are defined based on ASVs, each genome would be split into 5 OTUs.
Conversely, if the copies were clustered using a 3\% distance threshold
all 10 copies would cluster into the same OTU. The goal of this study
was to quantify the risk of splitting a single genome into multiple bins
and the risk of lumping together different bacterial species into the
same bin.

To investigate the variation in the number of copies of the 16S rRNA
gene per genome as well as the intragenomic variation among copies of
the 16S rRNA gene, I obtained reference 16S rRNA sequences from the rrn
copy number database (rrnDB; CITATION). Among the \textbf{4,774} species
represented in the rrnDB there were \textbf{15,614} genomes. The median
number of rrn operson per species ranged between \textbf{1}
(e.g.~\textbf{\emph{Mycobacterium tuberculosis}}) and \textbf{19}
(\textbf{\emph{Metabacillus litoralis}}) copies of the rrn operon. As
the number of copies of the operon in a genome increased, the number of
variants of the 16S rRNA gene in each genome also increased
(\textbf{FIGURE}). On average, there were \textbf{1} variants per copy
of the full length 16S rRNA gene and an average of \textbf{0.26},
\textbf{0.33}, and \textbf{0.27} variants when considering the V4,
V3-V4, and V4-V5 regions of the gene, respectively. Although a species
tended to have a consistent number of 16S rRNA gene copies per genome,
the number of total variants increased with the number of genomes that
were sampled (\textbf{FIGURE}). For example, \textbf{\emph{Mycobacterium
tuberculosis}} generally only had \textbf{1} copy of the gene per
genome, but across the \textbf{180} genomes that have been sequenced
there were \textbf{11} versions of the gene. Similarly, a \emph{E. coli}
genome typically had \textbf{7} copies of the 16S rRNA gene with between
\textbf{6} and \textbf{10} distinct full length sequences per genome.
Across the \textbf{958} \emph{E. coli} genomes that have been sequenced,
there were \textbf{1,013} different variants of the gene. These
observations highlight the risk of selecting a threshold for defining
units of inference that is too narrow because it is possible to split a
single genome into multiple units.

@@

To provide a more nuanced approach to selecting a threshold, it would be
useful to to quantify the sensitivity and specificity of characterizing
bacterial species using OTUs defined at different thresholds. I created
confusion matrices for multiple regions of the 16S rRNA gene: true
positives were those cases where two ASVs were joined in the same OTU
and the same species; true negatives were those cases where two ASVs
from different OTUs came from different species; false positives were
those ASVs that joined the same OTU, but were from different species;
and false negatives were those ASVs that joined different OTUs, but were
from the same species. By calculating the sensitivity and specificity
for each threshold and each region of the 16S rRNA gene, I was able to
constuct a receiver operator characteristic curve (ROC). Because the ROC
curve represents a range of possible thresholds and sensitivities and
specificities, I used two metrics to select the best threshold for
defining an OTU. First, I identified the thresholds where the
sensitivity and specificity were most similar to each other. For this
criterion, the best distance thresholds were \textbf{6.0}\% (V1-V9),
\textbf{4.5}\% (V4), \textbf{5.5}\% (V3-V4), and \textbf{4.0}\% (V4-V5).
Second, I identified the distance threshold that resulted in the point
on the ROC curve that was closest to perfect classification. For this
criterion, the best distance thresholds were \textbf{5.5}\% (V1-V9),
\textbf{3.5}\% (V4), \textbf{4.5}\% (V3-V4), and \textbf{3.5}\% (V4-V5).
Surprisingly, these analyses revealed that thresholds near 3\% distance
balance the risks of splitting genomes into separate OTUs and lumping
species into the same OTU.
